---
title: "Draft Report"
authors:
  - name: Mckenna Cooper
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Sarah Culhane
    affiliation: Colorado State University
    roles: writing
    corresponding: false
  - name: Archer Goodman
    affiliation: Colorado State University
    roles: writing
    corresponding: false
execute:
  echo: true
  
bibiography: references.bib
---
# Rivers and Lakes Bring the Boys to the Yard: Assessing the Effect of Riparian Zones on Taxonomic Richness

## Introduction
  Throughout the mythologies of ancient civilizations, rivers are often portrayed as sacred lifelines and mystical forces that breathe life into the land. From the Nile of Egypt to the Ganges in India, rivers shaped the very foundation of human development. The route a river carves through a landscape not only guides the path of its water, but also dictates the types of vegetation that grow alongside it, the diversity of animals that depend on it, and the feasibility of long-term human settlement. The majority of the world’s great cities were born beside rivers, not merely for commerce, which could be achieved from any coast, but for one elemental advantage: rivers carry fresh water. Drinkable water. Or at least, they used to.
  In today’s world, the demand for freshwater is greater than ever. Global population growth, expanding agricultural demands, urban sprawl, and climate change are placing unprecedented stress on aquatic ecosystems. Historically, human progress has been closely tied to our ability to manipulate the quantity of water available through irrigation systems, dams, aqueducts, and other technologies. But as environmental pressures grow, we are now forced to reckon with a second, equally critical dimension: water quality.
  Just as our ancestors learned to interpret the presence of certain plants and animals as signs of viable water sources, modern ecologists use biological indicators to evaluate the health of freshwater systems. Among the most widely used of these indicators are benthic macroinvertebrates; organisms that live on the bottom of rivers, lakes, and streams and include insects, crustaceans, mollusks, and worms. These organisms are uniquely valuable for biomonitoring due to their varied tolerance to pollutants, their limited mobility (which ties them to local conditions), and their life cycles, which span seasons or years.
  One particularly important group of macroinvertebrates used to indicate good water quality is the EPT taxa; Ephemeroptera (mayflies), Plecoptera (stoneflies), and Trichoptera (caddisflies). These taxa are generally sensitive to environmental disturbances, and their presence in high abundance is often correlated with clean, well-oxygenated water. In contrast, a dominance of more pollution-tolerant organisms such as oligochaete worms, leeches, and midge larvae (Chironomidae) is typically a sign of poor water quality and habitat degradation [@mullins1991macroinvertebrates].
  However, water quality is not influenced solely by what occurs within the streambed. The surrounding environment, particularly riparian zones, plays a critical role. Riparian zones are the transitional areas between terrestrial and aquatic ecosystems, typically encompassing the vegetation and soils adjacent to rivers and streams. Though narrow in spatial extent, riparian zones provide disproportionate ecological benefits. They act as natural buffers, filtering out sediments, nutrients, and pollutants before they reach the water body. They also stabilize stream banks, reduce erosion, moderate water temperature through shading, and provide crucial habitat for a wide range of terrestrial and aquatic organisms. As such, the integrity of riparian zones is often used as a secondary, yet equally important, indicator of watershed health [@sabo2005riparian].
  Unfortunately, riparian zones are frequently the first casualties of human development. Agricultural expansion, urbanization, logging, and road construction often involve the removal or fragmentation of riparian vegetation. When riparian zones are degraded, their ability to support both terrestrial and aquatic biodiversity diminishes, and the systems they protect become more vulnerable to pollution, sedimentation, and thermal stress. Consequently, monitoring the type and abundance of riparian vegetation can yield valuable insights into the overall ecological condition of a watershed, and by extension, its capacity to support a diverse and healthy macroinvertebrate community.
  With these connections in mind, our project seeks to investigate the relationship between riparian zone condition and freshwater invertebrate diversity. Specifically, we aim to evaluate whether the composition and abundance of riparian plant cover is associated with the composition and abundance of benthic macroinvertebrates, with an emphasis on EPT taxa as indicators of ecosystem health. By exploring these patterns, we hope to contribute to a better understanding of how human land use and vegetation management practices may impact water quality through both direct and indirect ecological pathways.
  To conduct this analysis, we will use data from the Klamath Network Lakes Monitoring Data Package (2013–2019), which includes biological and ecological monitoring data collected across three national parks: Crater Lake, Lassen Volcanic, and Redwood National Parks. From this dataset, we will focus on three specific tables: InvertSpecies.csv, which documents the abundance of macroinvertebrates at various sampling plots, RiparianCover.csv, which describes the vegetation characteristics of corresponding riparian zones, and WaterQuality.csv which includes data on physical and chemical water parameters at each sampling plot. Integrating this dataset will allow us to account for potential covariates or confounding variables, such as dissolved oxygen, pH, or turbidity, that may influence macroinvertebrate populations independently of vegetation conditions.
  Using this data, we will assign a standardized “vegetation score” to each plot based on the richness and type of plant cover recorded. This score will serve as a proxy for riparian zone quality. We will then analyze the relationship between this vegetation score and macroinvertebrate diversity, particularly the presence of sensitive EPT taxa, using statistical methods in RStudio. To determine the strength and direction of relationships between variables, we will conduct correlation tests (Pearson or Spearman, depending on normality assumptions). To model the predictive influence of vegetation on macroinvertebrate abundance, we will use linear regression analysis. These statistical approaches will be complemented by clear and informative data visualizations created with ggplot2, including scatterplots, box plots, and bar graphs to communicate trends and variation across sites. This holistic approach will allow us to better interpret the ecological relationships we uncover and provide more robust insights into freshwater health within these protected park systems.

## Data Exploration
  All data for this project comes from the Klamath Lakes Monitoring Data Package 2013–2019, originally compiled by the U.S. Department of the Interior and now publicly available through Data.com. This dataset includes ecological monitoring records from Crater Lake National Park (NP), Lassen Volcanic NP, and Redwood National and State Parks (NSP). However, for the purposes of our study, we will exclusively use data collected at Crater Lake National Park, narrowing the geographic focus to improve consistency and data quality.
  The three files central to our analysis are InvertSpecies.csv, RiparianCover.csv, and WaterQuality.csv. These files contain complementary data and share key identifying columns, including Lake_Code, Start_Date, and Park_Name—which will allow us to merge the datasets and compare biological, physical, and vegetative conditions at each sampled water body.
  The InvertSpecies file contains records of benthic macroinvertebrate communities sampled in lakes and ponds. It includes fields such as Taxon, Order, Count, and Life_Stage. For this study, we will focus on the orders Ephemeroptera (mayflies), Plecoptera (stoneflies), and Trichoptera (caddisflies), which are collectively known as EPT taxa. These macroinvertebrates are widely recognized as indicators of healthy freshwater ecosystems, as they are generally sensitive to pollution and habitat degradation.
  The RiparianCover file provides plot-level data on vegetation type and structure in riparian zones, such as Big_trees_ArealCover, Woody_shrubs_saplings_GroundCover, Canopy_Type, and Barren_AreaCcover. These variables will help us estimate the ecological integrity and potential human impact on surrounding shorelines. Each plot corresponds to a specific lake, allowing us to associate vegetation conditions with macroinvertebrate presence in those water bodies.
  The WaterQuality file offers insight into the abiotic conditions of each lake, including key measurements like Temperature_C, pH, Turbidity, Dissolved_oxygen, and Specific_conductivity. These parameters may contribute to or constrain macroinvertebrate diversity and could help contextualize any patterns observed between vegetation and aquatic biodiversity.
  While all three files have a relatively large number of rows (RiparianCover.csv: 1,680; InvertSpecies.csv: 3,299), some fields contain missing values, particularly with respect to sampling dates and environmental measurements. To account for this, we will filter for data collected in 2018, a year with strong representation across datasets, and group observations by plot. This will allow us to match vegetation conditions with corresponding macroinvertebrate counts and water quality measurements.
  From RiparianCover.csv, we are particularly interested in a subset of vegetation metrics that reflect both habitat structure and disturbance: Big_trees_ArealCover, Woody_shrubs_saplings_GroundCover, and Barren_ArealCover. These will form the basis of a vegetation condition score used later in the analysis. Meanwhile, from InvertSpecies.csv, we will extract insect counts and filter specifically for EPT taxa, which we define as “significantly present” when observed in quantities of at least 10 individuals per plot.
  Lastly, WaterQuality.csv will be used to examine physical and chemical characteristics of each lake. These variables will not only support broader ecological interpretations but also offer an additional layer of explanatory context for understanding macroinvertebrate patterns.
  By combining these three datasets with consistent identifiers and a focused temporal and spatial scope (Crater Lake, 2018), we can begin exploring the relationships among riparian cover, water quality, and macroinvertebrate biodiversity with clarity and coherence.

## Data Visualizations
Data Preparation
```{r}
# Load libraries
library(dplyr)
library(ggplot2)

# Load datasets
invert_species <- read.csv("InvertSpecies.csv")
riparian_cover <- read.csv("RiparianCover.csv")
water_quality <- read.csv("WaterQuality.csv")

# Filter for Crater Lake and 2018
invert_species_filtered <- invert_species %>%
  filter(Park_Name == "Crater Lake National Park" & Start_Date >= "2018-01-01" & Start_Date <= "2018-12-31")

riparian_cover_filtered <- riparian_cover %>%
  filter(Park_Name == "Crater Lake National Park" & Start_Date >= "2018-01-01" & Start_Date <= "2018-12-31")

water_quality_filtered <- water_quality %>%
  filter(Park_Name == "Crater Lake National Park" & Start_Date >= "2018-01-01" & Start_Date <= "2018-12-31")

# Merge datasets
merged_data <- invert_species_filtered %>%
  inner_join(riparian_cover_filtered, by = c("Lake_Code", "Park_Name", "Start_Date")) %>%
  inner_join(water_quality_filtered, by = c("Lake_Code", "Park_Name", "Start_Date"))
```
**Note: Although it is stated above and in Preliminary Methods that we would only be evaluating Carter Lake National Park in the year 2018, this was not feasible for the drafted visualizations. We had to pivot to a different proposal idea due to the elimination of data this past weekend, and need more time to narrow down how to visualize with these specific data sets.
1.
```{r}
library(ggplot2)
library(dplyr)

invert_species <- read.csv("InvertSpecies.csv")

invert_species %>%
  filter(
    Park_Name == "Crater Lake National Park",
    Start_Date >= "2013-01-01", Start_Date <= "2019-12-31",
    !is.na(Order),
    !is.na(Count)
  ) %>%
  group_by(Order) %>%
  summarize(Total_Count = sum(Count, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Order, -Total_Count), y = Total_Count, fill = Order)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Total Invertebrate Count by Order (Crater Lake, 2013-2019)",
    x = "Invertebrate Order",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

2. 
```{r}
# Load necessary libraries
library(ggplot2)
library(reshape2)

riparian_cover <- read.csv("RiparianCover.csv")

# Create a sample dataset
data <- data.frame(
  Park_Name = c("Redwood National and State Parks", "Lassen Volcanic National Park", "Crater Lake National Park"),
  Big_Trees = c(20, 15, 30),
  Small_Trees = c(10, 20, 15),
  Understory_Veg = c(30, 25, 20),
  Woody_Shrubs_Saplings = c(15, 10, 25),
  Herb_Grass_Forb = c(25, 30, 10)
)

# Melt the data for ggplot
data_melted <- melt(data, id.vars = "Park_Name", 
                    variable.name = "Vegetation_Type", 
                    value.name = "Areal_Cover")

# Plot stacked bar chart with tilted axis labels
ggplot(data_melted, aes(x = Park_Name, y = Areal_Cover, fill = Vegetation_Type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Riparian Vegetation Areal Cover", 
       x = "Park Name", 
       y = "Areal Cover (%)", 
       fill = "Vegetation Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Preliminary Methods
  To begin our analysis, we will first join the three datasets—InvertSpecies.csv, RiparianCover.csv, and WaterQuality.csv—using shared identifying columns: Lake_Code, Park_Name, and Start_Date. We will then filter the merged dataset to include only Crater Lake National Park and data collected in 2018, to maintain consistency in time and location.
  Next, we will divide our dataset into two groups of lakes based on the presence or absence of sensitive macroinvertebrate orders: Ephemeroptera, Plecoptera, and Trichoptera (collectively referred to as EPT taxa). These taxa serve as bioindicators of high water quality and intact riparian zones. A lake will be categorized as "EPT-present" if at least 10 individuals from any of the three orders are found. Lakes without EPT taxa or with fewer than 10 individuals will be categorized as "EPT-absent."
  Once we have established these two groups, we will assess potential relationships between riparian vegetation characteristics and the presence of EPT taxa. Key vegetation variables from RiparianCover.csv will include:
Big_trees_ArealCover


Woody_shrubs_saplings_GroundCover


Barren_ArealCover


  We will explore creating a vegetation integrity index by standardizing and summing Big_trees_ArealCover, Woody_shrubs_saplings_GroundCover, and inverse Barren_ArealCover scores.
We will also consider select abiotic water quality variables from WaterQuality.csv—such as Temperature_C, pH, and Dissolved_oxygen—to help contextualize our findings or identify confounding influences.
For statistical analysis, we will choose appropriate tests based on the type of variables being compared:

 - For categorical comparisons (e.g., EPT presence vs. vegetation category), we may use a Chi-squared test.


 - For numerical comparisons (e.g., EPT presence vs. percent vegetation cover), we will consider a T-test or ANOVA.


 - If relationships appear linear, we may explore correlation coefficients or linear regression models to assess strength and direction.


  To support our findings, we will use ggplot2 in R to create clear and informative visualizations—such as boxplots, bar charts, or scatterplots—highlighting any significant trends or differences between the EPT-present and EPT-absent lake groups.
Finally, missing data and uneven sampling will be carefully handled—either through exclusion of incomplete rows or imputation strategies, depending on the nature and extent of the gaps.

# Lab 9: Project Updates (Results and Discussion)

## Methods
  Our analysis focused on assessing the relationship between riparian vegetation characteristics and the diversity of freshwater macroinvertebrates, especially EPT taxa (Ephemeroptera, Plecoptera, Trichoptera), in Lassen Volcanic National Park for the years 2013 and 2019.
  We began by loading three datasets: InvertSpecies.csv, RiparianCover.csv, and WaterQuality.csv, filtering each to include only observations from Lassen Volcanic National Park in the target years. Using shared identifiers (Lake_Code, Start_Date, Park_Name), we merged the datasets for each year to create two comprehensive data frames, one for 2013 and one for 2019.
  To evaluate invertebrate community health, we calculated an EPT Score, defined as the ratio of EPT taxa counts to non-EPT taxa counts for each sampling date. Riparian vegetation quality was proxied through the categorical variable Barren_AreaCcover, which we converted into a numerical estimate of vegetative cover.
  We selected the 2013 dataset to build and evaluate a random forest regression model predicting EPT Score using water-quality and vegetation-related predictors: temperature in celsius, pH, dissolved oxygen, and estimated vegetative cover. Data was split into training and testing sets using initial_split(), and a cross-validation strategy (vfold_cv) was applied. We tuned model hyperparameters using tune_grid() and finalized the model based on RMSE performance.
To visualize patterns and support interpretation, we generated several plots using ggplot2, including:
Bar plots comparing insect order abundances by year


Boxplots of water quality metrics across years


A correlation heatmap to assess relationships between water quality parameters, EPT scores, and vegetation


A barren cover comparison plot between 2013 and 2019


All analyses were conducted in RStudio using the tidymodels, dplyr, and ggplot2 packages. 

## Drafting the Results
### Data Prep & Modeling: 
```{r}
#Loading Libraries
library(ggplot2)
library(dplyr)
library(tidymodels)
library(parsnip)
```

```{r}
#Load data
invert_species <- read.csv("InvertSpecies.csv")
riparian_cover <- read.csv("RiparianCover.csv")
water_quality <- read.csv("WaterQuality.csv")

# Filter for Lassen Volcanic National Park and our year of 2013
invert_species_2013 <- invert_species %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           (Start_Date >= as.Date("2013-01-01") & 
           Start_Date <= as.Date("2013-12-31")) |
           (Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31")))

riparian_cover_2013 <- riparian_cover %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           (Start_Date >= as.Date("2013-01-01") & 
           Start_Date <= as.Date("2013-12-31")) | 
           (Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31")))

water_quality_2013 <- water_quality %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           (Start_Date >= as.Date("2013-01-01") & 
           Start_Date <= as.Date("2013-12-31")) |
           (Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31")))

# Merge datasets
merged_data_2013 <- invert_species_2013 %>%
  inner_join(riparian_cover_2013, by = c("Lake_Code", "Park_Name", "Start_Date")) %>%
  inner_join(water_quality_2013, by = c("Lake_Code", "Park_Name", "Start_Date"))

#glimpse(merged_data_2013)

#Selecting our data for 2019
invert_species_2019 <- invert_species %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31"))

riparian_cover_2019 <- riparian_cover %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31"))

water_quality_2019 <- water_quality %>%
  filter(Park_Name == "Lassen Volcanic National Park" & 
           Start_Date >= as.Date("2019-01-01") & 
           Start_Date <= as.Date("2019-12-31"))

# Merge datasets
merged_data_2019 <- invert_species_2019 %>%
  inner_join(riparian_cover_2019, by = c("Lake_Code", "Park_Name", "Start_Date")) %>%
  inner_join(water_quality_2019, by = c("Lake_Code", "Park_Name", "Start_Date"))

#glimpse(merged_data_2019)
```

```{r}
#setting up some useful metrics, limiting data to stuff we need
melded_data_2013 <- merged_data_2013 |>
  mutate(EPT_taxa = ifelse(Order == "Ephemeroptera" |
                             Order == "Plecoptera" |
                             Order == "Trichoptera", 
                           "Yes", "No"),
         Veg_Cover = case_when(
           Barren_AreaCcover == "sparse (<10%)" ~ 0.05,
           Barren_AreaCcover == "moderate (10-40%)" ~ 0.25,
           Barren_AreaCcover == "heavy (40-75%)" ~ 0.57,
           Barren_AreaCcover == "very heavy (>75%)" ~ 0.87)) |>
  summarize(Unit_Code, 
            Park_Name, 
            Lake_Code, 
            Start_Date, 
            Order, 
            Count, 
            EPT_taxa, 
            Veg_Cover, 
            Temperature_C,
            Dissolved_oxygen, 
            pH, 
            Depth_m, 
            Turbidity, 
            )

melded_data_2013 <- melded_data_2013 |>
  group_by(Start_Date, Unit_Code) |>
  mutate(
    Yes_Total = sum(Count[EPT_taxa == "Yes"], na.rm = TRUE),
    No_Total = sum(Count[EPT_taxa == "No"], na.rm = TRUE),
    EPT_Score = Yes_Total / No_Total
  ) |>
  ungroup()

glimpse(melded_data_2013)

#and the same for 2019
melded_data_2019 <- merged_data_2019 |>
  mutate(EPT_taxa = ifelse(Order == "Ephemeroptera" |
                             Order == "Plecoptera" |
                             Order == "Trichoptera", 
                           "Yes", "No")) |>
  summarize(Unit_Code, 
            Park_Name, 
            Lake_Code, 
            Start_Date, 
            Order, 
            Count, 
            EPT_taxa, 
            Barren_AreaCcover, 
            Temperature_C, 
            Dissolved_oxygen, 
            pH, 
            Depth_m, 
            Turbidity, 
            )

melded_data_2019 <- melded_data_2019 |>
  group_by(Start_Date, Unit_Code) |>
  mutate(
    Yes_Total = sum(Count[EPT_taxa == "Yes"], na.rm = TRUE),
    No_Total = sum(Count[EPT_taxa == "No"], na.rm = TRUE),
    EPT_Score = Yes_Total / No_Total
  ) |>
  ungroup()
#glimpse(melded_data_2013)
```

```{r}
#Conduct a test. 
set.seed(15379)
#used my old robotics team's number for the seed, easy for me to remember

#We're gonna do this with just 2013 data for right now
#giving my data splits catchy names 
banana_split <- initial_split(melded_data_2013, prop = 0.8)
#Chocolate > choochoo > train > training data
choco_choochoo <- training(banana_split)
#vanilla > vanilla ice > mic check "testing testing" > testing data
vanilla_ice_mic <- testing(banana_split)

#glimpse(choco_choochoo)

#specifying a recipe
sundae_recipe <- recipe(EPT_Score ~ 
                          Start_Date +
                          Veg_Cover + 
                          Temperature_C + 
                          Dissolved_oxygen + 
                          pH + 
                          Depth_m + 
                          Turbidity, 
                        data = choco_choochoo) |>
  step_normalize(all_numeric_predictors())

#specifying a model
icecream_model <- rand_forest(mtry = tune(), trees = 250, min_n = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

#creating workflow
icecream_scoop <- workflow() |>
  add_model(icecream_model) |>
  add_recipe(sundae_recipe)

#testing the model with cross_validation
#validating the process for making a new icecream flavor
validated_process <- vfold_cv(vanilla_ice_mic, v = 5)
#tuning the model
salt_to_taste <- tune_grid(
  icecream_scoop,
  resamples = validated_process,
  grid = 10
)

#selecting the best model
#choosing the best formulas for the new icecream flavor
best_flavor <- select_best(salt_to_taste, metric = "rmse")

#finalizing the model
#selecting the winning flavor
final_flavor <- finalize_model(icecream_model, best_flavor)

#workflow with finalized model
#designing packaging for new flavor i guess?
final_flav_package <- icecream_scoop |>
  finalize_workflow(best_flavor)

#fitting the workflow to the training data
#shipping the new flavor to the stores
icecream_retail <- final_flav_package |>
  fit(data = choco_choochoo)

#making predictions
#seeing how the market reacts to the new flavor
public_tastes <- predict(icecream_retail, new_data = vanilla_ice_mic) |>
  bind_cols(vanilla_ice_mic)

#evaluating model performance
#Leaving a review of the ice cream flavor i guess
parlor_review <- public_tastes |>
  metrics(truth = EPT_Score, estimate = .pred)

#visualising model performance
ggplot(public_tastes, aes(x = EPT_Score, y = .pred)) + 
  geom_point(alpha = 0.5) + 
  geom_abline(linetype = "dashed", color = "blue") + 
  labs(x = "Actual", y = "Predicted", title = "Predicted vs Actual")

```
### For Taxa Order
```{r}
# Summarize 2013 data
insect_order_counts_2013 <- melded_data_2013 %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  group_by(Order) %>%
  summarise(Total_Count = sum(Count, na.rm = TRUE)) %>%
  mutate(Year = 2013)

# Summarize 2019 data
insect_order_counts_2019 <- melded_data_2019 %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  group_by(Order) %>%
  summarise(Total_Count = sum(Count, na.rm = TRUE)) %>%
  mutate(Year = 2019)

# Combine both
insect_order_counts_combined <- bind_rows(insect_order_counts_2013, insect_order_counts_2019)

# Plot
ggplot(insect_order_counts_combined, aes(x = reorder(Order, -Total_Count), y = Total_Count, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Comparison of Insect Counts by Order: 2013 vs 2019",
    x = "Insect Order",
    y = "Total Count",
    fill = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### For Barren Cover
```{r}
# Add year column to each dataset
barren_2013 <- merged_data_2013 %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  select(Barren_AreaCcover) %>%
  mutate(Year = 2013)

barren_2019 <- merged_data_2019 %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  select(Barren_AreaCcover) %>%
  mutate(Year = 2019)

# Combine both
barren_combined <- bind_rows(barren_2013, barren_2019)

# Plot
ggplot(barren_combined, aes(x = Barren_AreaCcover, fill = factor(Year))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Barren Area Cover in Lassen Volcanic National Park (2013 vs 2019)",
    x = "Barren Area Cover Category",
    y = "Number of Observations",
    fill = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### For Water Quality Box Plots
```{r}
# Combine data, reshape, and filter
water_quality_box <- bind_rows(
  melded_data_2013 %>% mutate(Year = "2013"),
  melded_data_2019 %>% mutate(Year = "2019")
) %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  select(Temperature_C, Dissolved_oxygen, pH, Year) %>%
  drop_na() %>%
  pivot_longer(cols = c(Temperature_C, Dissolved_oxygen, pH),
               names_to = "Metric", values_to = "Value")

# Create box plot
ggplot(water_quality_box, aes(x = Year, y = Value, fill = Year)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black") +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    title = "Box Plots of Water Quality Metrics (Lassen Volcanic National Park)",
    x = "Year",
    y = "Value"
  ) +
  scale_fill_manual(values = c("2013" = "blue", "2019" = "orange")) +
  theme_minimal() +
  theme(text = element_text(size = 13), legend.position = "none")

```
### For Correlation Heat Map
```{r}
# Combine 2013 and 2019 data, filter, and select relevant columns
cor_data_full <- bind_rows(
  melded_data_2013 %>% mutate(Year = "2013"),
  melded_data_2019 %>% mutate(Year = "2019")
) %>%
  filter(Park_Name == "Lassen Volcanic National Park") %>%
  select(Temperature_C, Dissolved_oxygen, pH, EPT_Score, Veg_Cover) %>%
  drop_na()

# Compute correlation matrix
cor_matrix_full <- cor(cor_data_full, use = "complete.obs")

# Reshape for heatmap
library(reshape2)
cor_melt_full <- melt(cor_matrix_full)

# Plot heatmap
ggplot(cor_melt_full, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "red", high = "blue", mid = "white",
    midpoint = 0, limit = c(-1, 1),
    name = "Correlation"
  ) +
  labs(
    title = "Correlation Heatmap: Water Quality, EPT Score, and Barren Cover (Lassen Volcanic National Park)",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Discussion
  Results suggest that there was an overall decline in macroinvertebrate taxa richness across all orders between 2013 and 2019 in water bodies within Lassen Volcanic National Park. The strong positive correlation between pH, dissolved oxygen, and EPT_score suggests that the decline in pH and dissolved oxygen present in those water bodies may have been partially responsible for the decline in taxa richness. Vegetation present in riparian zones within the park seemed to have increased, according to our analysis, and veg_cover and EPT_score seem to have a negative correlation. This suggests that as vegetation cover increased, taxa richness decreased, which goes against our hypothesis that more riparian cover would lead to high taxa richness. The cause of this is unknown and warrants further research. Going forward, more focused research and data collection needs to be done to find the source of pollution in Lassen Volcanic National Park so that measures can be taken to combat pollution at the source and improve the water quality of water bodies within the park. Limitations of our analysis include the fact that there was a lot of cleaning that had to be done in our dataset, and there was a disproportionate amount of data for each of the national parks and years that data was collected, with some parks and years having significantly more data than others. Crater Lake National Park especially had less data than both Volcanic National Park and Redwood National State Park.

